Directionality in Kilo-ton Scale Scintillating Neutrino Detectors

### 2013-07-15 Direction.tex

%
\author{C.~Aberle}\ucla
\author{A.~Elagin}\chicago
\author{L.~Winslow}\ucla

@@

\section{Light Production in Liquid Scintillators}
Molecules vibrate...index of refraction = Cerenkov.

>>>>

%
\author{C.~Aberle}\ucla
\author{A.~Elagin}\chicago
\author{M.~Wetstein}\chicago
\author{L.~Winslow}\ucla

@@

\section{Light Production in Liquid Scintillators}
Liquid scintillators are cocktails of aromatic hydrocarbons. When charged particles move through a scintillator, the molecules are excited, predominantly the the non-localized electrons in the $\pi$-bonds of the phenyl groups. Vibrational and rotational modes of the molecules are turned into heat within picoseconds through collisions with other molecules.  Within $\sim$10 picoseconds, the $\pi$-electrons de-excite to the first excited state from higher levels through radiationless transitions. The first excited state de-excites through photon emission. There are two characteristic times for this de-excitation depending if the singlet state or the triplet state was excited.  The singlet state will de-excite within nanoseconds while the triplet state de-excites on the order of 10's or 100's of nanoseconds. These two processes are fluorescence and phosphorescence respectively. The exact time constants for these processes are determined by the composition of the scintillator.

The molecules in liquid scintillators are not isolated. Radiationless processes transfer energy between the molecules. The probability of energy transfer between molecules increases as the overlap between the molecule's absorption and emission spectra increase. The absorption and emission spectra overlap at some level for all molecules. Therefore, if there is only one type of molecule in the scintillator cocktail, the light output would be reduced due to inefficiencies in the energy transfer through multiple absorption and reemission processes. Aromatic solutes or fluors are added to the primary solvent to shift the wavelength of the photons to longer wavelengths where the scintillator is more transparent. This wavelength-shifting is also used to match the quantum efficiency as a function of wavelength for the photodetectors being used. One typical scintillator mixture uses pseudo-cumene as the solvent with 1-5~g/L of PPO as the fluor. This mixture has a peak emission at about 400~nm where photomultiplier tubes (PMTs) are most sensitive and the pseudo-cumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons isotropically per MeV of deposited energy. Although less abundant, if a particle is moving faster than the speed of light in the medium then Cerenkov light will be produced as well. This light is emitted in a cone pointed in the direction of the particles trajectory and with a continuous spectrum wavelengths weighted toward shorter wavelengths but extending well into the red,
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and $\beta$ is the velocity of the incoming particle. The Cerenkov light produced at wavelengths shorter than the absorption cutoff of the scintillator will be absorbed and re-emitted as isotropic light, but wavelengths longer than this cutoff will propagate across the detector retaining their directional information, roughly 60 photons per MeV assuming a 400~nm absorption cutoff. These undisturbed Cerenkov photons will have timing determined by the group velocity in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) + dn(\lambda)/dlog(\lambda)}.
\end{equation}
These longer wavelengths Cerenkov photons arrive in principle before the scintillation light which is slowed by both the scintillation processes and the shorter wavelengths involved. Thus, with sufficient timing resolution and sensitivity to longer wavelength, it should be possible to separate the directional Cerenkov light and the isotropic scintillation light in time and then reconstruct the direction of the initial particle.
